---
title: "Exercise 2"
output: html_document
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "p03_outputs")
  })
---

# Exercise introduction

This is Exercise 2 in Part 1 of the course.

The purpose of the exercise is to give students more familiarity with `R`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basics of R

Use the `sudoku` package to view a data matrix.

```{r sudoku}
activatePkgs('sudoku')
library(sudoku)
puz <- fetchSudokuUK()
puz
class(puz)
```

## Basic data aggregations

Going through some basic data creation and aggregations. This chunk includes vectors as well as a dataframe.

```{r dataCreationAndAggs}
x <- c(2, 3, 4, 5, 4, 7, 9, 6, 7, 2)
x

mean(x)
median(x)

y <- c(3, 4, 6, 5, 4, 7, 6, 5, 7, 2)
z <- c(1, 1, 1, 1, 1, 2, 2, 2, 2, 2)

df <- data.frame(x, y, z)

df

```

## Plotting

`R`'s default `plot` command with continuous data is a scatterplot.

```{r scatterplot}
plot(y ~ x, data = df)
```

## Modifying plot parameters

Modifying parameters will modify the output from the `plot` function.

```{r modifiedScatterplot}
plot(y ~ x, data = df, las = 1, xlab = 'x axis', ylab = 'y axis', cex.lab = 1.5, pch = 2)
```

## Creating a simple linear model

Use the function `lm` to create a simple linear model. The `summary` function called upon the linear model will provide information about the residuals, the coefficients for the model, and even a basic t-test.

```{r simpleLinearRegression}
reg <- lm(y ~ x, data = df)

summary(reg)
```

## Plotting the linear model

Use the `abline` function to after creating a plot in order to add the linear model to the scatterplot.

```{r plottingLm}
plot(y ~ x, data = df, las = 1, xlab = 'x-axis', ylab = 'y-axis', cex.lab = 1.5, pch = 2)
abline(reg, col = 'blue')
```

## Adjusting plot limits

If the plot is too short or narrow, one can modify the parameters `ylim` and `xlim` to modify the plot limits.

```{r adjustLimits}
plot(y ~ x, data = df, las = 1, xlab = 'x-axis', ylab = 'y-axis', cex.lab = 1.5, pch = 2, xlim = c(0, 9))
abline(reg, col = 'blue')
```

## Plotting continuous \~ categorical data

If you have continuous data in two or more groups/categories, then a boxplot is a standard way to visualize the results. This is the default output of `R`'s `plot` function if the 'independent' variable in formula is of type `factor`.

```{r boxplot}
df$w <- as.factor(df$z)

plot(y ~ w, data = df)
```

## `gplots` package

There are many packages for plotting. One is `gplots` which has the function `plotmeans` for plotting means as well as 95% confidence intervals.

```{r meansPlot}
activatePkgs('gplots')
plotmeans(y ~ w, data = df, las = 1, ylim = c(2, 9)
          , connect = F
          , barcol = 'black', cex = 2, pch = 16)
```

The `connect` argument must be set to `FALSE` or else the means will be connected with a line.

```{r connectedMeans}
activatePkgs('gplots')
plotmeans(y ~ w, data = df, las = 1, ylim = c(2, 9)
          # , connect = F
          , barcol = 'black', cex = 2, pch = 16)
```

## Aggregations

One can call the `mean` function on multiple vectors in a single call by using the `aggregate` function with the `by` parameter.

```{r aggregations}
aggregate(x = df$y, by = list(df$w), FUN = mean)
```

## Welch's two-sample t-test

Using `t.test` will also provide the means in addition to the t-test results.

```{r tTest}
t.test(y ~ w, data = df)
```

# Key learnings

-   Library `gplots` has a relatively convenient function `plotmeans`\
-   One can use `aggregate` with `by` and `FUN` parameters to accomplish aggregations on group levels

# Unresolved questions

-   Which type of t-test is used when calling `summary` on a linear model object?\
-   How to specify different 'flavors' of t-tests with the `t.test` function? For example, one-sample vs. two-sample, paired vs. unpaired, homoscedasticity assumption fulfilled or not (Levene's adjustment)?
