---
title: "Replace with title"
author: "Replace with name"
date: "Replace with date"
output: 
  word_document: 
    fig_height: 7
    fig_width: 7
---


```{r, message=FALSE, warning=FALSE, include=FALSE}
library(Rcmdr) #this chunk start RCommander 
library(ggplot2) #this cunks starts ggplot2
#if you have used other packeges you may need to include them here
```


```{r, message=FALSE, warning=FALSE, include=FALSE}
#include a chunk that opens your indata
```


## Introduction

This is where you write the background and the aim with the assigment.


## Method

Brief explanation of methods, and a description of the work-flow.

## Result

Include the important result from your analyses. Have a short informative running text explaining the results.

Add scripts in R chunks. Additional chunks can be included with the "insert"" button 


```{r}

```

```{r}

```



## Discussion

Discuss your findings and make conclusions. Be sure to answer questions. 

## References

Include references if appropriate
ex. homepages that you fetch data from, articles from scientific journals, etc. 
