---
title: "Exercise 1"
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "p03_outputs")
  })
output:
  html_document:
    toc: true
---

# Exercise introduction

This is Exercise 1 in Part 4 of the course.

The purpose of the exercise is to cover ANOVA and Kruskal Wallis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1: Different costs between different housing options in Stockholm  

### Step 1: Import data   

```{r dataImport}
housing <- readr::read_tsv('../p02_inputs/Cost_month.txt') %>% tibble()
```

### Step 2: One-way ANOVA  

```{r owAnova}
library(mvtnorm, pos=70)
library(survival, pos=70)
library(TH.data, pos=70)
library(multcomp, pos=70)
library(abind, pos=74)

housing %<>%
    mutate(Housing = as.factor(Housing))

costModel <- aov(Costpermonth_SEK ~ Housing, data=housing)

oldpar <- par(oma=c(0,0,3,0), mfrow=c(2,2))
plot(costModel)
par(oldpar)
```

```{r boxcoxTransform}
library(MASS)
bc <- boxcox(Costpermonth_SEK ~ Housing, data = housing)
lambda <- bc$x[which.max(bc$y)]
bcCostModel <- aov(((Costpermonth_SEK ^ lambda - 1) / lambda) ~ Housing, data = housing)
oldpar <- par(oma=c(0,0,3,0), mfrow=c(2,2))
plot(bcCostModel)
par(oldpar)
```



```{r anovaPosthocs}
summary(bcCostModel)
summary(costModel)
library(RcmdrMisc)

with(housing, numSummary(Costpermonth_SEK, groups=Housing, statistics=c("mean", "sd")))
local({
  .Pairs <- glht(bcCostModel, linfct = mcp(Housing = "Tukey"))
  print(summary(.Pairs)) # pairwise tests
  print(confint(.Pairs, level=0.95)) # confidence intervals
  print(cld(.Pairs, level=0.05)) # compact letter display
  old.oma <- par(oma=c(0, 5, 0, 0))
  plot(confint(.Pairs))
  par(old.oma)
})



local({
  .Pairs <- glht(costModel, linfct = mcp(Housing = "Tukey"))
  print(summary(.Pairs)) # pairwise tests
  print(confint(.Pairs, level=0.95)) # confidence intervals
  print(cld(.Pairs, level=0.05)) # compact letter display
  old.oma <- par(oma=c(0, 5, 0, 0))
  plot(confint(.Pairs))
  par(old.oma)
})

```

## Part 2:  Different stress factors effects on heart rate  

I did this exercise for another course, so am skipping it here.


## Part 3: Chlorophyll-α concentrations (μg/l) in the different basins of the Baltic Sea  

```{r loadChData}
(chlor <- readr::read_tsv('../p02_inputs/Chla.txt') %>% tibble())
```


```{r chlrAno}
chlor %<>%
    mutate(Basin = as.factor(Basin)
           , LARGE_basin = as.factor(LARGE_basin)
           , Chlorophyll_a = as.numeric(str_replace(Chlorophyll_a, ',', '.')))

bcChlor <- boxcox(Chlorophyll_a ~ LARGE_basin, data = chlor)
lambda <- bc$x[which.max(bcChlor$y)]
bcChlorAov <- aov(((Chlorophyll_a ^ lambda - 1) / lambda) ~ LARGE_basin, data = chlor)
oldpar <- par(oma=c(0,0,3,0), mfrow=c(2,2))
plot(bcChlorAov)
par(oldpar)
```

```{r aovPwChlor}
summary(bcChlorAov)

with(chlor, numSummary(Chlorophyll_a, groups=LARGE_basin, statistics=c("mean", "sd")))
local({
  .Pairs <- glht(bcChlorAov, linfct = mcp(LARGE_basin = "Tukey"))
  print(summary(.Pairs)) # pairwise tests
  print(confint(.Pairs, level=0.95)) # confidence intervals
  print(cld(.Pairs, level=0.05)) # compact letter display
  old.oma <- par(oma=c(0, 5, 0, 0))
  plot(confint(.Pairs))
  par(old.oma)
})
```

Using Kruskal-Wallis as an alternative.  
```{r kw}
# install.packages('pgirmess')
library(pgirmess)
library(PMCMRplus)

attach(chlor)
pgirmess::kruskalmc(Chlorophyll_a, LARGE_basin, probs = 0.05, cont=NULL)#resp=your response variable, cat= is your categorical varible
kruskal_posthoc <- PMCMRplus::kwAllPairsNemenyiTest(x = Chlorophyll_a, g = LARGE_basin, dist = "Tukey")
summary(kruskal_posthoc)
detach(chlor)
```
```{r boxplotChlor}
Boxplot(Chlorophyll_a ~ LARGE_basin, data = chlor, id=list(method="y"), outline= FALSE)
```


# Key learnings

-   Function `aov` and `lm` both take formula syntax, and both accept dependent variables expressed as complex expressions, permitting box-cox transformations. However, their output is slightly different in that a `lm` that uses an independent factor variable will by default split that variable into k-1 estimates where k is the number of factor levels, whereas `aov` retains it as a simple variable. This can be seen by the call of `summary(<model>)`  
-   The box-cox transformation is easy to implement, but does not necessarily result in lower p-values even if the model is a better overall fit.  
-   Can use either `pgirmess::kruskalmc` or `PMCMRpluss::kwAllPairsNemenyiTest` to do Kruskal-Wallis tests with pairwise post-hoc comparisons.  

# Unresolved questions

-   How to use `mcp` with alternatives other than 'Tukey'? For example, if I wanted an 'fdr' post-hoc correction?  
-   Why does function `Boxplot` still include some labeled outliers if set `outline = FALSE`?  
